Agile Formality: A Mole of Software Verification Practices

### 2010-06-30 papers/ModelBasedTestCaseGenerationDataStreamProtocol/paper.tex
problems with citations

% problems would benefit form agile and formal mix, characterize the problem
The two worlds seem irreconcilable, nevertheless, there exist problems that would greatly benefit from both.
The problem needs to be at least instable in the requirements definition, constrained by deadlines that cannot be postponed, but also characterized by artifacts that needs to be formally specified and verified.
The complex problem addressed in the case study fits these needs.

\todo{describe, ``Formal Versus Agile: Survival Of the Fittes'', blending advantages and potentials}

@@

% what needs to be considered: TDD
The most common software verification practices in agile methodologies need to be considered as well when tring to integrate formal and agile methodologies, since they often are a fundamental part of the development process. 
The most significant verification practice is Test Driven Development (TDD), TDD is a software development technique, originally defined in Extreme Programming (XP) methodology, characterized by writing unit tests before actually writing the code.
Actually, the general technique, consisting in writing the tests before the code, is used at high abstraction levels as well, since user stories (the requirements) are translated into test cases (the functional tests) before implementation.
The test driven apprach is a cornerstone of XP, it is so popular that have been adopted in other agile methodologies as well; also, it is consider a good software development technique to be used on its own, regardless of the surrounfing development process.
It is to be noted that, depending on the abstraction level, the test driven approach can introduce activities cycles that can be as long as a complete iteration (functional tests), or as short as a couple of minutes (small increment on a unit test).

% previous very related work
Some tentative attempts to reconcile formal and agile methodologies have been developed.
The recurrent approach of the attempts consists in removing the hand made test suites, substituing them with different kinds of automated verification based on the system formal specification; the automated verification can be as simple as enabling run time verification of assertions or automatically generating a test suite based on the system properties, as complex as actually proving the whole system behaviour through static model and code checkers or theorem provers.

\todo{describe and reference ``An Agile Formal Development Methodology'', perfect match}

\todo{describe and reference ``Formal Extreme (and Extremely Formal) Programming'', perfect match}

\todo{describe and reference ``XFM: An Incremental Methodology for Developing Formal Models'', perfect match}

% previous work: a common approach
Unfortunately, the agile and formal reconciliation attempts that have been defined so far are not entirely satisfing.
As already mentioned, in all these attempts, the hand made test suites are automatically generated using the system formal specification, or completely replaced by theorem provers or static checkers. 
The requisite to enable these approaches is, at least, a specification of the entire system under development to enable simple verification techniques as run time verification (obtained by assertions and design by contract precondition, postconditions and invariants) \todo{reference Meyer} and test suites automatic generation \todo{references, JML, Eiffel, \ldots}.
At most, a complete and sound system specification, usually supported by an appropriate formal language (to specify model and properties) and a constrained programming language, to be able to actually prove properties about the system under development \todo{references, ESC/2, \ldots}.

% previous work: why is not satisfying
The methodologies are sound, but they share some common problems: the loops between development artifacts and related activities (code, tests, design, refactoring), especially the closed and tight ones, typical of TDD, are completely lost. 
This is not a minor issue because it is believed that these loops are the inner engine of many agile methodologies, imposing to the development process an iterative and incremental pace \todo{find meaningful references, I know there are some}.
Beside possibly disrupting some of the agile methodologies inner mechanincs, only the verification purpose of the tests is mantained, that is, the documentation and requirement specification purposes are lost and need to be moved on top of different artifacts, usually the formal specification.
Finally a complete system formal specification is needed to apply these methodologies, but this requirement is hard to fulfill since the effort required to write a complete formal specification in most cases is much greater than writing a suite of tests.

% what we want to achieve
Thus, the objective set is different: we want to maintain an higly iterative and incremental development process, we want to connect development artifacts and activities (code, requirements, tests, formal specifications, refactorings, run-time verification, automatically generated tests, \ldots) through activities cycles similar to what we can find in agile methodologies (especially in XP), we want to be able not to formally specify the complete system in order to cope with constrained resources and

@@

\todo{quick introduction}

\todo{define what we are looking at: verification, cycles and dependencies between artifacts}

\todo{the cycles in SCRUM and XP}

\todo{TDD}

@@

The Java Modeling Language (JML) is the de facto standard specification language for describing the behavioral properties of Java programs.  JML is a rich behavioral interface specification language (BISL) and thus focuses on the modular specification of classes and objects in which inheritance is used solely for behavioral subtyping~\cite{X}.   JML includes standard specification constructs like assumptions, assertions, axioms, and pre- and postconditions, framing clauses, and invariants.  It also includes a rich model-based specification layer that includes mathematical models, data refinements~\cite{X}, datagroups~\cite{X}, and many other advanced object-oriented specification features~\cite{advancedJMLpaper}.  Many tools, ranging from compilers to static checkers to full-functional verification systems support the JML language~\cite{STTTpaper}.

JML is sometimes used in a Design-by-Contract style, where a specification is written from scratch, reusing existing APIs that have specifications of their own, and then an implementation is written conforming to that specification~\cite{DBCjml,DBC}.  The implementation is checked by whatever means appropriate using some subset of the aforementioned tools.  At other times an existing piece of implemented and tested code is annotated with specifications after-the-fact, a process we call Contracting-the-Design.

The two tools used most frequently to check the correctness of implementations are the JML Runtime Assertion Checker (RAC) and the Extended Static Checker for Java, version 2 (ESC/Java2)~\cite{RAC,ESC}.  The former compiles executable assertions into runtime checks, thereby generating inline test code executed at runtime.  Using RAC is familiar for developers as unit testing is en vogue at the moment.  The latter tool performs automated modular semantic static analysis to attempt to prove that the program under analysis does not misbehave (i.e., crash in a variety of ways) and conforms to its specification (lightweight functional correctness).  As ESC/Java2 is fully automated, it is not sound, nor does it cover the whole Java and JML languages, and is thus not complete, but there is support for warning the user about both of these situations~\cite{soundness,completeness}.

@@

% =====================================================================
\section{A Real And Complex Case Study: Rapid Development In Small Scale Hardware Software Co-design}
\label{sec:a_real_and_complex_case_study}

The \textbf{UCD CASL SenseTile System} is a large-scale, general-purpose sensor system to be installed at the University College Dublin in Dublin, Ireland.  
Our facility aims to provide a scalable and extensible infrastructure for performing in-depth investigation of both the specific and general issues in large-scale sensor networking.
This system integrates a sensor platform, a datastore, and a compute farm.  
The sensor platform is a custom-designed, but inexpensive, sensor platform (called the \ST) paired with general-purpose small-scale compute nodes, including everything from low-power PDAs to powerful touchscreen-based portable computers.  
There are over a dozen sensors packaged on the \ST itself, the board is expandable as well, new sensors can be added to it easily.

The case study is focused in building the sensor board and its software driver side by side.
Because of hard time constraints, and the initial unavailability of the custom board, we need to progress concurrently with all the development tasks, which consist of: (1) specification of the communication protocol; (2) physical board specification; (3) physical board; (4) board embedded software; (5) communication protocol software driver; (6) development of software simulators of the board.

The development of the physical board, along with its embedded software (development tasks 2, 3, 4), is done by an external manufacturer, thus it is not directly taken into account here. The specification of the communication protocol (development task 1) is a joint effort, while the remainang tasks (5, 6) can be developed in isolation.

\todo{illustrate dependencies: 3<>2<>4<>1, 1>5, 1>6}

\todo{explain 3<>2<>4<>1: difficulties encountered}

\todo{explain 1>5, 1>6: clean separation, but 1 is unstable}

There exist various approaches in literature addressing this kind of development constraints (rapid development in hardware software
co-design), but they are all focused on big-scale systems.
The hardware software formal co-design methodologies usually have several common development artifacts\cite{Slomka2000,Hoffman2001}: an high level specification of the system, a translation (refinement) of the high level specification to low level ones (hardware and software counterparts), the possibility to generate software code from the low level specifications, a hardware simulator capable of simulating an hardware component based on hardware specification, the hardware component developed.

When considering the development of the device and of its software driver as separate entities, that possibly have to be developed concurrently, the existing approaches are similar to the ones seen in the case of hardware software co-design\cite{Valderrama1995,Siegmund2002,Ryzhyk2009}: they all focus on a specification that aims to be as complete as possible.

As it has been already pointed out, a specification as complete as possible is not feasible nor convenient: we want to enhance the protocol specification incrementally, and we are not able to provide a complete specification.

The protocol specification is divided in various (thin) layers, and each of the layer has been specified with a different lightweight approach: some of the layers have been specified formally, while others have been specified in a less formal way. 
The formal specifications are used to check at run time the correctness of the \SB, the driver and the simulators.  
The formal specifications are used to automatically generate a test suite to verify the driver and the simulators, and eventually to verify the
specification itself.  
Hand made test suites are used to verify the driver and the simulators as well.  
The generated and hand made test suites are used together, complementing each other.

The approach proved itself effective, at the point that, when the board has been finally released and delivered, only one small defect has been found
in the driver.
The errors and the not agreed changes introduced in the board and in the protocol by the hardware supplier have been discovered immediately, with no additional effort, while the simulators were able to feed with correct data the rest of the system.

\subsection{Data stream and protocol description}
\label{sec:data_stream_and_protocol_description}

The \SB protocol is asynchronous and packet based, the packets have a fixed length.  
The protocol is separated in multiple layers to ease its formal description, since mixing up different abstraction levels can lead to overly complex and unmanageable descriptions, especially when dealing with formal ones.  
The layers identified in protocol are the following: (1) packet byte structure: the internal representation of the packet; (2) packet info structure: the meaningful fields contained in the packets; (3) single packet rules: the content acceptable values, and how they influence each other; (4) packet sequence rules: the content acceptable values, based on values of previously received packets; (5) packet input output asynchronous rules: the content acceptable
  values and reaction constraints on output packets, based on previously transmitted input packets.

We focus our attention to the first three layers, considering only the output sensor data packet.
The packet byte and info structures of the output sensor data packet reflect the board capabilities and built in sensors. 
Essentially a single packet has to accommodate various types of data: fast, medium and slow data rate streams, together with metada describing the sensors and the board state.

The internal structure of the package is strictly fixed, even if the data streams have different data rates. 
This choice originates from a trade-off between two constraints: the efficiency of the transmission (redundant data to be minimized) and the complexity to build a package (complexity to be minimized, since the FPGA capabilities on the \STSB are limited). 
A packet is internally composed by frames. 
The frames (82 frames in each packet) accommodate data from the fast and medium data rate streams (there are 4 fast data rate stream and 8 medium data rate stream channels) and theirs associated metadata.

Fast, medium and slow data rates have not a common multiple; a packet, to be capable to accommodate three different rates outputs in a fixed structure, relies in two forms of redundancies, for slow and medium data rates:

@@

\subsection{\ldots}

\todo{adapt old paper}

@@

% old paper
\chapter{OLD, to be reused}

>>>>

% problems would benefit form agile and formal mix, characterize the problem
The two worlds seem irreconcilable, nevertheless, there exist problems that would greatly benefit from both~\cite{Black2009}.
The problems' class we are interested in need to have unstable requirements, to be constrained by deadlines that cannot be postponed, but also to be characterized by artifacts that need to be formally specified and verified.

\subsection{Formal and agile integration}
\label{subsec:formal_and_agile_integration}

@@

% what needs to be considered: TDD
The most common software verification practices in agile methodologies need to be considered, when tring to integrate formal and agile methodologies, since they are a fundamental part of the development process. 
The most significant verification practice is Test Driven Development (TDD)~\cite{Beck2003}, TDD is a software development technique, originally defined in Extreme Programming (XP)]\cite{Beck2004} methodology, characterized by writing unit tests before writing the code.
Actually, the general technique, consisting in writing the tests before the code, is used at high abstraction levels as well, since user stories (the requirements) are translated into test cases (the functional tests) before implementation.
The test driven apprach is a cornerstone of XP, it is so popular that have been adopted in other agile methodologies as well; also, it is considered a good software development technique to be used on its own, regardless of the surrounfing development process.
It is to be noted that, depending on the abstraction level, the test driven approach can introduce activities cycles that can be as long as a complete iteration (functional tests), or as short as a couple of minutes (small increment on a unit test).

% previous related work
Some tentative attempts to reconcile formal and agile methodologies have been developed.
The recurrent approach of the attempts consists in removing the hand made test suites, substituing them with different kinds of automated verification based on the system formal specification; the automated verification can be as simple as enabling run time verification of assertions or automatically generating a test suite based on the system properties, or as complex as actually proving the whole system behaviour through static model and code checkers or theorem provers.

% a quick analysis of the related works
Eleftherakis and Cowling in~\cite{Eleftherakis2003} propose XFun, an iterative and incremental process to generate X-Machines models, an extension of Finite State Machines. 
The development process impose a complete specification of the system, and its verification is done exclusively using the tools provided by the X-Machines formal methodology. 
The development process proposed in~\cite{Herranz2003b} by Herranz and Moreno-Navarro is quite similar; they propose an iterative process to model a system using SLAM-SL, an object-oriented specification language, and its tools. 
Some XP practices are fully adopted or considered compatible, as pair-programming, iterative and incremental development, system refactoring; while tests are completely replaced by the verification tools provided by the SLAM suite.
A different formal methodology, but same approach, can be found also in~\cite{Suhaib2005} where XFM methodology is explained.

% previous work: a common approach
As already mentioned, in all these attempts, the hand made test suites are automatically generated using the system formal specification, or completely replaced by theorem provers or static checkers. 
The requisite to enable these approaches is, at least, a specification of the entire system under development to enable simple verification techniques as run time verification (obtained by assertions and design by contract~\cite{Meyer1997} precondition, postconditions and invariants) and test suites automatic generation~\cite{Cheon2002,Cheon2004,Cheon2005}.
At most, a complete and sound system specification, usually supported by an appropriate formal language (to specify model and properties) and a constrained programming language, to be able to actually prove properties about the system under development~/cite{CatanoHuisman02,DetlefsNelsonSaxe2005,KiniryCok04}.

% previous work: why is not satisfying
The approach is simple: the verification, documentation and requirement specification purposes of the tests are moved on to the formal specifications, and a complete formal specification of the system under development is needed to apply these methodologies.
The problem is that this requirement is hard to fulfill, since the effort required to write a complete formal specification of the system, in most real world complex cases, is much greater than writing a suite of tests.
If a complete formal specification is not built, all the previous methodologies share a common problem: the parts not formally specified are not verified.
The parts not formally specified can obviously be verified with the traditional methods, but the loops between development artifacts and related activities (code, tests, design, refactoring) are not specified or completely lost.
The loops and relations between the different development artifacts are the inner engine of many agile methodologies, they impose an iterative and incremental pace to the development process.

% what we want to achieve
The objectives we want to achive are the following ones: to be able to formally specify only parts os the system, in order to cope with constrained resources and unstable requirements; to draft an higly iterative and incremental development process that blends formal and agile practices.

% how we want to achive it
The main specific problem, by now never addressed, is how to connect development artifacts of the two worlds together: we will define activity cycles, similar to what is found in agile methodologies (especially in XP), to solve this problem.
We will show how tests can drive the formal specification, how the formal specifications can drive tests and code development, how hand made and automated tests can coexist and support each other, how the unspecified parts of the system can be incrementally specified.
That is, we will show how to blend, and not compose, formal and agile verification practices: a Mole of verification practices.

Mole practices are applied in the context of a real world case study that matches the problem requirements previously detailed: unstable requirements and development artifacts that need to be formally specified and verified.
The case study fits these needs.

\subsection{A Real And Complex Case Study: Rapid Development In Small Scale Hardware Software Co-design}
\label{subsec:a_real_and_complex_case_study}

% SenseTile introduction
The \textbf{UCD CASL SenseTile System} is a large-scale, general-purpose sensor system installed at the University College Dublin in Dublin, Ireland.
Our facility provide a scalable and extensible infrastructure for performing in-depth investigation of both the specific and general issues in large-scale sensor networking.
This system integrates a sensor platform, a datastore, and a compute farm.  
The sensor platform is a custom-designed, but inexpensive, sensor platform (called the \ST) paired with general-purpose small-scale compute nodes, including everything from low-power PDAs to powerful touchscreen-based portable computers.  
There are over a dozen sensors packaged on the \ST itself, the board is expandable as well, new sensors can be added to it easily.

% case study: the sensor board development tasks
The case study is focused in building the sensor board and its software driver side by side.
Because of hard time constraints, and the initial unavailability of the custom board, we need to progress concurrently with all the development tasks, which consist of: (1) specification of the communication protocol; (2) physical board specification; (3) physical board; (4) board embedded software; (5) communication protocol software driver; (6) development of software simulators of the board.

% what we have to do, what we outsourced, and the dependencies between tasks
The development of the physical board, along with its embedded software (development tasks 2, 3, 4), is carried out by an external manufacturer, thus it is not directly taken into account here. 
The specification of the communication protocol (development task 1) is a joint effort, while the remainang tasks (5, 6) can be developed in isolation.
The dependencies are straighforward: the specification of the communication protocol (1) is the most stable element, but it still depends on the sensor board (2, 3, 4): big changes in the latter could affect the former.
The driver and the simulators depends directly on the communication protocol.

% how the problem is usually managed
There exist various approaches in literature addressing this kind of development constraints (rapid development in hardware software
co-design), but they are all focused on big-scale systems.
The hardware software formal co-design methodologies usually have several common development artifacts\cite{Slomka2000,Hoffman2001}: an high level specification of the system, a translation (refinement) of the high level specification to low level ones (hardware and software counterparts), the possibility to generate software code from the low level specifications, a hardware simulator capable of simulating an hardware component based on hardware specification, the hardware component developed.
When considering the development of the device and of its software driver as separate entities, that possibly have to be developed concurrently, the existing approaches are similar to the ones seen in the case of hardware software co-design\cite{Valderrama1995,Siegmund2002,Ryzhyk2009}: they all focus on a specification that aims to be as complete as possible.

% complete specification is not feasible
A specification as complete as possible is not feasible nor convenient in this case: the protocol specification have to be enhanced incrementally, and its complete specification cannot be provided.
This makes the problem an ideal candidate to test the Mole practices.

% formal methodology constraints
An appriate formal methodology needs to be chosen; it must be able to support a specification that is built incrementally and accompanied by tools that allow the verification of the system properties at run time.
All the formal methods that require a complete specification to be used effectivelt must be avoided.
The formal language should also be able to automatically generate tests.

% the chosen formal methodology
Being Java the language chosen for the implementation, the Java Modeling Language (JML)~\cite{advancedJMLpaper}, accompanied by the JML Runtime Assertion Checker suite of tools~\cite{RAC}\footnote{\myhref{http://sourceforge.net/projects/jmlspecs/}{Common JML tools}} has been selected.
JML and RAC tools provide a run time checker, a test generator, and do not require a complete specification to be used effectively.

\subsection{\todo{paper structure title}}

\todo{explain paper structure}

@@

Agile methodologies are all based on a highly iterative and incremental process.
They share a common approach on team management, customer relation, simplification and removal of not needed artifacts and activities and importance of working solutions and customers satisfaction.
The agile manifesto~\cite{Beck2001} summarizes the philosphy and principles shared by all agile methodologies.

In the present context, focused on how to merge agile and formal practices, we are mainly interested in the verification practices used in agile methodologies.
The agile manifesto does not introduce any specific techniques to be used in the development process, it describes agile methodologies from a very high abstraction level.
Nevertheless most of the agile methodologies share a similar approach to artifacts verification.
Functional test suites, to be used as precise requirements or user stories definition, are used in DSDM~\cite{Stapleton1997}, XP and Crystal Clear~\cite{Cockburn2004}, while unit tests and TDD are mentioned, whether not completely accepted, in most of the agile methodologies defined so far.

High abstraction level test suites, functional tests, describe requirements, and actually bind together requirements documentation and system behaviour, the development cycle is one iteration long. 
Low abstraction level test suites, unit tests, describe a module behaviour, and actually bind together code documentation, code and the contract imposed on the module, the development cycle is usually less than one day.

Refactoring is possible and feasible because of the safety nets provided by the test suites, used as regression test suites.
Continuous refactoring~\cite{Fowler1999} enables an interative and incremental development, since the system architecture and design, whether explicitly specified or not, needs to be cleaned up, after implementation, and modified, to accomodate new requirements, services and modules.
The automated test suites guarantees that the functionalities and behaviour already implemented in the code are not compromised by the system refactorings.

@@

The Java Modeling Language (JML)\footnote{\myhref{http://www.cs.ucf.edu/~leavens/JML/}{The Java Modeling Language (JML)}} is the de facto standard specification language for describing the behavioral properties of Java programs.  
JML is a rich behavioral interface specification language (BISL) and thus focuses on the modular specification of classes and objects in which inheritance is used solely for behavioral subtyping~\cite{X}. 
JML includes standard specification constructs like assumptions, assertions, axioms, and pre- and postconditions, framing clauses, and invariants. 
It also includes a rich model-based specification layer that includes mathematical models, data refinements~\cite{X}, datagroups~\cite{X}, and many other advanced object-oriented specification features~\cite{advancedJMLpaper}. 
Many tools, ranging from compilers to static checkers to full-functional verification systems support the JML language~\cite{STTTpaper}.

JML is sometimes used in a Design-by-Contract style, where a specification is written from scratch, reusing existing APIs that have specifications of their own, and then an implementation is written conforming to that specification~\cite{DBCjml,DBC}.  
The implementation is checked by whatever means appropriate using some subset of the aforementioned tools.  
At other times an existing piece of implemented and tested code is annotated with specifications after-the-fact, a process called Contracting-the-Design.

The two tools used most frequently to check the correctness of implementations are the JML Runtime Assertion Checker (RAC) and the Extended Static Checker for Java, version 2 (ESC/Java2)~\cite{RAC,BurdyEtal05-STTT,ESC}.
The former compiles executable assertions into runtime checks.  
The latter tool performs automated modular semantic static analysis to attempt to prove that the program under analysis does not misbehave (i.e., crash in a variety of ways) and conforms to its specification (lightweight functional correctness). 
As ESC/Java2 is fully automated, it is not sound, nor does it cover the whole Java and JML languages, and is thus not complete, but there is support for warning the user about both of these situations~\cite{soundness,completeness}.

@@

% =====================================================================
\section{\todo{the case study}}
\label{sec:___}

\subsection{Data stream and protocol description}
\label{subsec:data_stream_and_protocol_description}

% protocol layers
The \SB protocol is asynchronous and packet based, the packets have a fixed length.  
The protocol is separated in multiple layers to ease its implementation in the driver, since mixing up different abstraction levels can lead to overly complex and unmanageable code.  
The layers identified in protocol are the following: (1) packet byte structure: the internal representation of the packet; (2) packet info structure: the meaningful fields contained in the packets; (3) single packet rules: the content acceptable values, and how they influence each other; (4) packet sequence rules: the content acceptable values, based on values of previously received packets; (5) packet input output asynchronous rules: the content acceptable values and reaction constraints on output packets, based on previously transmitted input packets.
For space reason, the first three layers are analyzed in the case study, considering only the output sensor data packet.

% packet structure
The packet byte and info structures of the output sensor data packet reflect the board capabilities and built in sensors. 
A single packet has to accommodate various types of data: fast, medium and slow data rate streams, together with metada describing the sensors and the board state.
The internal structure of the packet is strictly fixed, even if the data streams have different data rates. 
This choice originates from a trade-off between two constraints: the efficiency of the transmission (redundant data to be minimized) and the complexity to build a packet (complexity to be minimized, since the processing capabilities in the \STSB are limited).

% packet internal structure: collection of frames
A packet is internally composed by frames. 
A frame (82 frames in each packet) accommodate data from the fast and medium data rate streams (there are 4 fast data rate stream and 8 medium data rate stream channels) and theirs associated metadata.
Fast, medium and slow data rates have not a common multiple; a packet, to be capable to accommodate outputs at different rates, in a fixed structure, relies in two forms of redundancies, for slow and medium data rates:

@@

\subsection{Data stream and protocol specification}
\label{subsec:data_stream_and_protocol_specification}

% protocol specification
The specifications of the protocol are distilled and refined incrementally.
The protocol is divided in various (thin) layers, and each of the layer has been verified with a different approach: some of the layers have been specified formally.

\paragraph*{Packet byte structure specification}

The unformal specification and verification is provided by a hand made test suite, composed of unit and integration tests; the tests specifies the behaviors of the implementation, which is capable of recognizing the proper packet structure in a binary data stream. 
The implementation is tricky: it checks for the sentinels repeatedly to be sufficiently confidant that the byte stream can be properly decomposed into a packet stream.

The specification can be used to verify a board simulator, but it cannot be used to check properties at run time.
Code and tests have been implemented using TDD.

\paragraph*{Packet info structure specification}

The specification is provided by hand made test suites and JML annatations. 
It has been built starting from hand made tests only, and incrementally adding tests and later on, JML annotations.

\todo{explain: test -> code, test + code -> formal specification, formal specification -> generated tests}

\paragraph*{Single packet rules}

The specification is provided exclusively by the formal specification with JML language.
It has been built starting from JML annoatations, supported by hand made tests when needed.

\todo{formal specification -> test + generated tests + code}

% why simulators are useful
Simulators are used to test parts of the system minimizing dependencies: a simulator can be used by upper layers, with no need to provide the functionalities of the lower layers. 
The simulators proved essential during development, since the physical board was not available.

% how many of them do we need?
Theoretically a separate simulator is needed for each protocol layer, to specify, develop and test each layer in isolation.  
But, because of limited resources and affort avilable, the layer structure was not matched by corresponding simulators.
Two simulators are available.

% simulators development
Simulators have been built concurrently with the specification, the driver implementation, and the tests.

% simulators description
The driver API is splitted in two abstraction layers: the general high level interfaces, that expose the main functionalities of the board and the main contents of the packets, and the lower level implementation that parse the data streams in and out the board, meant to translate the higher level instructions and data in properly formed packets.
The simulators are built according to these two abstraction layers.
The high level simulator implements the interfaces providing the methods to deal with an abstracted \SB: the simulator knows nothing
about the details of the real format of the binary data streams.
The low level simulator is capable to rebuild the \SB data streams: the in and out data streams are built exactly as the sensor board is expected to be able to parse or generate.

% a simple specification example
The JML specifications are of varying complexities.  
Some of them are rather simple, focusing on constraints that should hold when calling a method (the preconditions) and constraints on the return value (the very basic form of postconditions). 
In listing \ref{lst:jml_example_simple} a simple JML specification example is shown, the method \lil{getTemperature} is declared to be \lil{/*@ pure @*/}, which means that it cannot change the state of any object (a postcondition), the specification also constraints the return value with a lower and upper bound (another postcondition).

% a complex specification example
The complex specifications usually focus on properties regarding the behavior of a whole object.
In listing \ref{lst:jml_example_invariant} a complex invariant example is shown; an invariant is a property that has to be maintained during the life cycle of an object, more precisly, an invariant is assumed on entry and guaranteed on exit of each method of an object. 
The invariant is constraining the number of samples for each medium data rate streams: the total number of streams is \lil{Frame.ADC_CHANNELS}, the total number of frames is \lil{FRAMES}, the constant constraining the number of samples is \lil{FRAMES/Frame.ADC_CHANNELS+1}, meaning that the samples contained in a frame are fairly distributed on the channels.
The valid samples are counted parsing all the frames contained in a packet, selecting only the matching valid samples.  
A medium data rate stream sample is considered valid when \lil{isADCActive()} method returns \lil{true}.

\begin{lstlisting}[
  caption={A specification with JML annotation: a complex object invariant which constraints 
    the number of samples for each medium data rate stream channel.},
  label=lst:jml_example_invariant,
  float=htbp]
  /*@ 
      invariant ( 
        \forall int channel; 0 <= channel && channel < Frame.ADC_CHANNELS; ( 
          \num_of int i; 0<= i && i<(FRAMES-1); (
            (getFrame(i).isADCActive()) && 
            (getFrame(i).getADCChannel() == channel)
          ) 
        ) <= (FRAMES / Frame.ADC_CHANNELS + 1) 
      ); 
  @*/
\end{lstlisting}

\subsection{Test cases}
\label{subsec:test_cases}

% unit tests kinds
The unit test cases that verify the protocol driver are of two kinds: hand made unit tests and automatically generated unit tests based on JML specifications; they are built to be used together, they are complementary.

% unit tests effectiveness
The tests effectiveness is evaluated for each test suite; the evaluation is carried out in section \ref{subsec:test_cases_retrospectives}.
The test effectiveness evaluation considers three elements: effective results on piloting the real board (quantitative), code coverage
(quantitative), development help and usefulness (qualitative).

% the driver package structure
The package structure of the driver is shown in Figure \ref{fig:class_diagram_main}: two independent packages are defined (\lil{Stream} and \lil{Driver}).  
The packages are abstract, that is, they mainly contain abstractions; in Java language this is translated into a package which contains mainly abstract classes or interfaces.  
The dependency between the two abstract packages is not direct, it is realized through an implementation (\lil{StreamDriver}).
This is needed to maintain a high decoupling of the packages, and is the result of applying the dependency inversion principle~\cite{Martin1996}.

% the test package structure
The test package structure, which is shown in Figure \ref{fig:class_diagram_test} reflects and mimics the code structure; the tests for an abstract package are abstract, and implemented by the package that tests a corresponding system implementation. 
This is a well known test pattern (the Abstract Test Pattern~\cite{Thomas2004}) used to test that the contracts defined in the abstractions are respected in the several implementations.  
For instance, package \lil{DriverT} contains abstract tests for the abstractions of package \lil{Driver}, package \lil{StreamDriverT} inherits the abstractions contained in \lil{DriverT} and makes them concrete, to test the corresponding implementation \lil{StreamDriver}; package \lil{StreamDriverT} also contains stand alone tests written specifically for the implementation contained in \lil{StreamDriver}.

% the generated test package structure
The JML specifications are used to generate tests.
The resulting test package structure is shown in Figure \ref{fig:class_diagram_generatedtest}.  
The package dependencies are defined by the specifications; for instance, \lil{SimulatorDriverST} are generated tests, they test (hence use) the implementation \lil{SimulatorDriver}, and have been generated using the corresponding specification \lil{SimulatorDriverS}.  
The generated tests do not maintain the inheritance structure of the code and the specification.

% simulator roles in tests
The simulators are used in the hand made test suits as stubs.
The resulting structure is shown in Figure \ref{fig:class_diagram_streamdriver_test}.  
For instance, in Figure \ref{fig:class_diagram_streamdriver_test} the test suite \lil{StreamDriverT} is using the \lil{SimulatorStream} to properly simulate the \lil{Stream} used by \lil{StreamDriver}, which is the system under test.

% simulator importance
The use of simulators to enable unit testing was essential to test the \STSB driver, since the physical board was not available.
The simulators are used in both the hand made and the generated test suite, since both need a proper set of stubs to be executed.

\subsection{Test cases on the job}
\label{subsec:test_cases_on_the_job}

% test suites sizes
The hand made test suite is composed by over 140 tests, while the generated one is 100 times bigger, totaling over 14000 tests.
The big number of generated tests can be understood looking closely at how the generated tests are created by the JML framework. 
The generated tests explore the possible input combinations on public methods, combining the type data ranges that are specified for the test suite (see \cite{Cheon-Leavens02} for more details on how the JML framework works on generating unit tests). 
For instance, let's suppose we want to test a method \lil{m(int p1, int p2)} for a class \lil{C}; the data ranges specified for type \lil{int} are \lil{\{1,2,3\}} and for type \lil{C} are the instances \lil{\{c1,c2,c3,c4\}}. 
The JML framework generates $24 = 3 \times 3 \times 4$ tests, since they are the combination of the data ranges involved (the data range of the parameters and the data range of the type owning the method under test). 
Neverteless, not all the generated tests are executed.
A generated test is not executed, unless the preconditions of the called method are satisfied. 
Thus, depending on the preconditions specified for a method, the number of the generated tests that are actually executed in the test suite can be significantly smaller compared to the total number of generated tests.

% test execution time, values
Test suites execution time are different, the hand made test suite, with the run time assertion checking active, runs completely in less than one minute on an average PC, while the generated test suite runs in more than 20 minutes (the rough ratio is $ 1 : 60 $). 
Most of the time is spent on setting up the suite, only considering the real execution of the tests, with no suite setup, the execution time reduces to less than 4 minutes (the rough ratio is $ 1 : 10 $).

% test effectiveness
Code coverage metrics is used to compare in a quantitative way the effectiveness of the test suites. 
This is only an indicator, since it is accepted that code coverage alone cannot asses the quality of a test suite\cite{Marick1999,Chockler2006}.
Various coverage metrics exist, the test suites are analyzed with statement code coverage (one of the simplest types)\footnote{The tool used to obtain statement code coverage metrics is \myhref{http://emma.sourceforge.net/}{Emma}.}: statement coverage reports whether each statement has been encountered during execution. 
Statement code coverage is easy to implement, but lacks the capability to explore the control flow graph associated to the code, that is,
decision branches are not explored. 
The coverage results are reported in the Table \ref{tab:statement_code_coverage}.

\subsection{Retrospective on test cases effectiveness}
\label{subsec:test_cases_retrospectives}

% <100 coverage measures, utility methods
The coverage measures reported in Table \ref{tab:statement_code_coverage} do not reach 100\%. 
This is because of two different effects: non public utility methods, and abstract methods in interfaces.
The effect of non public methods (package Java visibility) is that these methods are taken into account as public and protected ones are, during code coverage calculation; these methods are not part of the interface, they are not meant to be called by the system, and they are only used in object initialization or in object setups performed during unit tests. 
In fact there are package and private code blocks that are not reachable by construction from public and protected methods, and this is a precise characteristic reflecting the code architecture, meant to ease class instances setup and configuration for testing purposes. 
The used statement coverage tool cannot be configured to distinguish these code blocks, so this effect is unavoidable. 
The effect is mainly seen in low \lil{SimulatorDriver} code coverage figures: the \lil{SimulatorDriver} package has many utility non public methods.

% <100 coverage measures, interfaces
Regarding interfaces, an interface contains no statements, so when an interface method is called, it is the method of the implementation class used that is actually covered.
This effect is visible in \lil{Driver} coverage figures: \lil{Driver} package contains interfaces (that are not counted at all) and some very simple real class (exceptions) that are not throughly tested. 
The net result is extremely low statement code coverage figures because the very simple classes are the only ones that are actually covered.

% code coverage interpretation
The statement code coverage figures show that the test suites do a good job on the most important package, the \lil{StreamDriver}. 
The generated test suite reaches almost $ 80 \% $, while the hand made tests have only a slightly lower coverage ratio: $ 76 \% $. 
The combination of test suites raise the coverage ratio to $ 88.5 \% $. 
This is a clear indication that one test suite is not completely overlapping the other: the intuition that test suites are not to be considered alternatives, but they need to be used together to achieve the higher benefits, is confirmed.
A similar indication is provided by \lil{SimulatorDriver} package as well.

% hand made unit test vs generated unit test
The qualitative difference of the two test suites can be understood observing how the suites are built. 
An hand made unit test usually requires objects initialization (it can involve several methods call), the call to the method under test, and the following assertions to verify the expected states change on the objects involved. 
A generated unit test has a more focused and limited scope: only one method call is performed in each test, it is difficult to explore complex behaviors. 
With a complete specification, and specifying large enough data ranges, the generated tests can achieve the same expressive power as the hand made ones; but the specification is not complete, and a large enough data ranges would take long to be developed.
The test suites give their best when used in a combined approach: the generated test suite, checking simpler and formalized behaviors, and the hand made, checking the more complex ones, whether formalized or not.

@@

\section{Extracts}

\paragraph*{limited resources}

% effect of the limited resources on the formal specifications
Hard constraints on driver delivery deadlines define the completeness level of the (\emph{just good enough}) specification.  
In \STSB driver JML specifies very simple behaviors (method oriented) on the whole driver implementation, and more complex behavior on the driver core interfaces.  
The software system parts not specified by the JML2 specification were specified using hand made unit tests.

\paragraph*{hand made unit tests}

% hand made unit tests purposes
Hand made unit tests specifies the behavior not specified by formal specification; nevertheless this is not the only purpose they have.
Hand made unit tests have been used to drive the development, since the Test Driven Development approach\cite{beck2003test} has been
followed during development.
